--- 
title: "Clinical Trials 4H"
author: "Rachel Oughton"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook:

    highlight: tango
    css: "style.css"
    config:
      toc:
        collapse: section
      bookdown:
      sharing:
        github: yes
        facebook: no
        twitter: no
  bookdown::pdf_book:
    highlight: tango
    classoption:
      - onecolumn
      - portrait
    extra_dependencies: 
      geometry: ["margin = 1in"]
      hyperref: ["breaklinks=true"]
    latex_engine: pdflatex
    toc: true
    keep_tex: yes
    citation_package: natbib
    df_print: kable
documentclass: book
classoption: openany
bibliography: [ct4h.bib]
biblio-style: apalike
always_allow_html: true
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "These notes mirror what we'll follow in lectures for Clinical Trials 4H. If you have any questions or notice any errors, please email me (Rachel Oughton)."
---

```{r setup, include = F}
knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# This line sometimes needs commenting out to first generate the html file
options(knitr.table.format = function() {
  if (knitr::is_latex_output())
    "latex" else "pipe"
 })
```

```{r, echo=F}
library(medicaldata)
library(dplyr)
library(ggplot2)
library(ggfortify)
library(gridExtra)
library(knitr)
library(kableExtra)
library(magick)
library(htmlTable)
library(magrittr)
library(rstatix)
library(survival)
library(mvtnorm)
library(randomizeR)
library(Minirand)
require(blockrand)
library(cvcrand)
#library(eefAnalytics)
library(shiny)
library(pROC)
library(parmsurvfit)
library(HSAUR3)
library(sjPlot)
library(lme4)
library(GGally)
library(gganimate)
library(gtsummary)
library(RColorBrewer)
library(jtools)
library(car)
library(ggsurvfit)
library(survminer)
```

# Welcome to Clinical Trials 4H! {-}

This page contains the notes for Clinical Trials IV 2024-25. As we progress through the course, more will appear. You can also download the PDF version (see the icon at the top left). If you notice any typos, mistakes or places that are unclear, please do let me know!

In case you're reading the PDF, the notes are best viewed in HTML (the link is [https://racheloughton.github.io/ct4/](https://racheloughton.github.io/ct4/)). The PDF will be up-to-date, but the formatting is designed for HTML.

## Practical details {-}

### Lectures {-}

Our lectures are 9am on Wednesdays in W007 (this is the Appleby Building, between Pysics and the main library), and 3pm on Fridays in PCL 048 (this is in the ground floor of the Palatine Centre). 

### Computer classes {-}

We have four computer practicals for this module. They are 1-2pm on the Mondays of weeks 13, 15, 17 and 19. These classes are in MCS 3098.

### Office Hour {-}

The office hour will be every Wednesday, 10-11am, in the Calman Learning Centre cafe. 

### Assessment {-}

This module is assessed through two equally weighted pieces of coursework. The first will be assigned on Wednesday 7th February (due 3rd March), the second on Wednesday 19th March (due 5th May).

There will also be two formative assignments during the course. More details on these to follow.

### Books and resources {-}

The main reference for the first half of the course is @matthews2006introduction. There are a couple of copies in the Bill Bryson Library. Some other books we will make use of are @hulley2013designing, @hayes2017cluster. You shouldn't need to use any of these books, but of course you're welcome to if you want to read further.

I will share links along the way for relevant resources like podcast episodes and articles, which I hope will help you to get a feel for the topic more generally.

## What to expect from this module {-}

Clinical Trials IV is somewhat different from the majority of statistics modules, because

  * Its main focus is on application
  * It is assessed purely through coursework.
  
This means that your experience of it might be different from what you're used to

  * We will cover quite a lot of different statistical methods (drawing on most of the 1H and 2H courses, and some 3H!) but not in great depth
  * There is no pressure to memorize anything - indeed, if you really were a trial statistician, you would definitely have access to the internet, various textbooks and even these notes (should they prove useful!).
  * There is an emphasis on understanding which method we use and why, and what it means. Hopefully this has been the case in some of your other modules too!
  
### What I expect from you {-}

Because we will be covering quite a lot of different areas within statistics, there may be some things that you haven't seen before (or can't remember very well). I will try my best to explain them as clearly as I can, but there isn't time to go into the nuts and bolts of everything we come across. Therefore, if you do feel a bit rusty on some area, you may need to read up on that a bit, so that you're happy with it. I am very happy to suggest resources from time to time, and you're welcome to come to the office hour to talk about such things.

This course is in its infancy, and so I would also really appreciate your feedback. I may not be able to address everything (or I may only be able to implement things for following years), but if I can act on it quickly then I will!
  





